---
title: "Sensitivity Analysis and Optimization"
author: "Florian Hartig"
date: "22 May 2016"
output: html_document
---


First, load the BT package

```{r}
library(BayesianTools)
```

# The Very Simple Ecosystem Model (VSEM)

We will use the Very Simple Ecosystem Model (VSEM), a model that is provided in the BayesianTools package. The VSEM is a 'toy' model designed to be very simple but yet bear some resemblance to deterministic processed based ecosystem models (PBMs) that are commonly used in forest modelling.

The model determines the accumulation of carbon in the plant and soil from the growth of the plant via photosynthesis and senescence to the soil which respires carbon back to the atmosphere.

Gross Primary Productivity (GPP) is calculated using a very simple light-use efficiency (LUE) formulation multiplied by light interception. Light interception is calculated via Beer's law with a constant light extinction coefficient operating on Leaf Area Index (LAI).

A parameter (GAMMA) determines the fraction of GPP that is autotrophic respiration. The Net Primary Productivity (NPP) is then allocated to above and below-ground vegetation via a fixed allocation fraction. Carbon is lost from the plant pools to a single soil pool via fixed turnover rates. Heterotrophic respiration in the soil is determined via a soil turnover rate. The model time-step is daily.

Parameters and state variables of the model are described in

```{r}
?VSEM
```

To get some data to drive the model, we use the VSEMcreatePAR() function to create a random radiation (PAR) time series. We visualize the time series using the plotTimeSeries() function from the BayesianTools package.


```{r}
PAR <- VSEMcreatePAR(1:1000)
plotTimeSeries(observed = PAR)
```


To run the model, we first need some parameters. The following function obtains a number of reference parameters, together with their upper and lower limits for calibration. To these we add an additional parameter for the standard deviation of the observation error that we will add to our data (error-sd). 

```{r}
refPars <- VSEMgetDefaults()
refPars[12,] <- c(0.1, 0.001, 0.5)
rownames(refPars)[12] <- "error-sd"
head(refPars)
```

Using the reference parameters, we create a model prediction (referenceData) and observations (obs).


```{r}
referenceData <- VSEM(refPars$best[1:11], PAR) 
obs <- referenceData + rnorm(length(referenceData), sd = (abs(referenceData) + 1E-7) * refPars$best[12])
oldpar <- par(mfrow = c(2,2))
for (i in 1:4) plotTimeSeries(observed = obs[,i], predicted = referenceData[,i], main = colnames(referenceData)[i])
```

# Sensitivity analysis

```{r}
parSel = c(1:6)

sensitivityTarget <- function(par){
  x = refPars$best
  x[parSel] = par
  predicted <- VSEM(x[1:11], PAR)
  return(sum(predicted[,2]))
}
```


Now a small trick - we'll use a special function from the BayesianTools package that is not intended for SA, but it's useful because this transforms the simple target into the structure that is expected by the sensitivity package.

```{r}
bayesianSetup <- createBayesianSetup(likelihood = sensitivityTarget, lower = refPars$lower[parSel], upper = refPars$upper[parSel], best = refPars$best[parSel], names = rownames(refPars)[parSel])
targetFunction <- bayesianSetup$posterior$density
sensitivityTarget(refPars$best[parSel])
```


```{r}
library(sensitivity)
```

Look at the helop of the package - there are a lot of functions available - here just an example

```{r}
morrisOut <- morris(model = targetFunction, factors = rownames(refPars[parSel, ]), r = 200, design = list(type = "oat", levels = 5, grid.jump = 3), binf = refPars$lower[parSel], bsup = refPars$upper[parSel], scale = TRUE)
par(mfrow=c(1,1))
plot(morrisOut)
```

\section{Calibrating the VSEM via optimization}

Let's create some observations with the reference paramters

```{r}
observation = VSEM(refPars$best[1:11], PAR)[,1]
```

Now we forget that we know the true parameters and try to find the parameters that fit best to this data.

Let's for the moment only change 2 parameters (parameters 1,3) to make it easy


```{r}
parSel = c(1,3)

optimizationTarget <- function(par){
  x = refPars$best
  x[parSel] = par
  predicted <- VSEM(x[1:11], PAR)[,1]
  return(sum(abs(predicted - observation)))
}


startvalue = refPars$lower[parSel]

optimResult = optim(startvalue, optimizationTarget)

optimResult
refPars$best[parSel]

```

At the moment we are only calibrating 2 parameters, i.e. parameters 1,3 - now take parameter 2 to the list and observe what happens?

