---
title: "Proposal"
author: "Florian Hartig"
date: "7 Jan 2016"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Put the title of your vignette here}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8](inputenc)
---


```{r, echo = F, message = F}
set.seed(123)
library(BayesianTools)
```



# Core classes 

## The Bayesiansetup


```{r}

likelihood <- function(x) dnorm(x, log = T)
prior <- createUniformPrior(0,5)
bayesianSetup <- createBayesianSetup(likelihood, prior)
str(bayesianSetup)

```



### Parallelization issues


Create reference data

```{r}
PAR <- VSEMcreatePAR()
plot(PAR)
pars   = c(0.5 , 1.5, 0.002,   0.4, 1440., 27370.)
referenceData <- VSEM(pars, PAR) 
referenceDataError <- referenceData #* rnorm(dim(referenceData)[1] * dim(referenceData)[2], mean = 1, sd = 0.1)
```

Likelihood

```{r}

lower = c(0.3,1,0.001, 0.2, 1000, 10000, 0.1)
upper = c(1, 3, 0.004, 1, 2000, 50000, 2)
prior = createUniformPrior(lower = lower, upper = upper)

likelihood <- function(x){
  pars = x[1:6]
  sd = x[7]
  predicted <- VSEM(pars, PAR)
  return(likelihoodIidNormal(predicted[,1], referenceDataError[,1], sd))
}

```

BayesianSetup 


```{r}
bayesianSetup = createBayesianSetup(likelihood, prior)

bayesianSetup$posterior$density(rbind(lower, lower))

```



## The proposalgenerator


```{r}

testMatrix = matrix(rep(c(0,0,0,0), 1000), ncol = 4)
testVector = c(0,0,0,0)


## Standard multivariate normal proposal generator

testGenerator <- createProposalGenerator(covariance = c(1,1,1,1), message = T)

methods(class = "proposalGenerator")
print(testGenerator)

testGenerator$returnProposal(testVector)
x <- testGenerator$returnProposalMatrix(testMatrix)
boxplot(x)

## Changing the covariance
testGenerator$covariance = diag(rep(100,4))
testGenerator <- testGenerator$updateProposalGenerator(testGenerator, message = T)

testGenerator$returnProposal(testVector)
x <- testGenerator$returnProposalMatrix(testMatrix)
boxplot(x)


## Changing the gibbs probabilities / probability to modify 1-n parameters

testGenerator$gibbsProbabilities = c(1,1,0,0)
testGenerator <- testGenerator$updateProposalGenerator(testGenerator)

testGenerator$returnProposal(testVector)
x <- testGenerator$returnProposalMatrix(testMatrix)
boxplot(x)


## Changing the gibbs weights / probability to pick each parameter

testGenerator$gibbsWeights = c(0.3,0.3,0.3,100)
testGenerator <- testGenerator$updateProposalGenerator(testGenerator)

testGenerator$returnProposal(testVector)
x <- testGenerator$returnProposalMatrix(testMatrix)
boxplot(x)


## Adding another function

otherFunction <- function(x) sample.int(10,1)

testGenerator <- createProposalGenerator(
                                            covariance = c(1,1,1), 
                                            otherDistribution = otherFunction, 
                                            otherDistributionLocation = c(0,0,0,1)
)

testGenerator$returnProposal(testVector)
x <- testGenerator$returnProposalMatrix(testMatrix)
boxplot(x)
table(x[,4])

```





# The Metropolis sampler 












## A more complicated problem, different options

Correlated, 3-d multivariate normal function

```{r}
ll <- generateTestDensityMultiNormal(sample = T)
pairs(ll(1000))
```

Setup

```{r}
ll <- generateTestDensityMultiNormal(sample = F)
bayesianSetup = createBayesianSetup(likelihood = ll, best = c(1,1,0))
```

Run the sampler with prior optimization

```{r}
testSampler <- mcmcSampler(bayesianSetup = bayesianSetup, optimize = T)
testSampler2 <- getSamples(testSampler, 10000, adapt = T)

plot(testSampler2)
correlationPlot(testSampler2)
marginalPlot(testSampler2)
```

```{r}

testSampler2 <- getSamples(testSampler, 10000, adapt = T, DRlevels = 2)

plot(testSampler2)
correlationPlot(testSampler2)
marginalPlot(testSampler2)

```








# The SMC sampler 




```{r, echo = F, message = F}
set.seed(123)
library(BayesianTools)
```


# Basic demonstration using Banana-shaped target function


### Sequential evaluation


```{r}
bayesianSetup <- createBayesianSetup(testDensityBanana, lower = c(-5,-5), upper = c(5,5))
system.time(finalParticles <- smcSampler(bayesianSetup, initialParticles = 1000, iterations = 5, resampling = T, adaptive = T))
```

### Plot result

```{r}
marginalPlot(finalParticles)
correlationPlot(finalParticles)
```


### Parallel evaluation with automatic selection of cores

n-1 of all n available cores will automatically be selected

```{r}
bayesianSetup <- createBayesianSetup(testDensityBanana, lower = c(-5,-5), upper = c(5,5), parallel = T)
system.time(finalParticles <- smcSampler(bayesianSetup, initialParticles = 5000, iterations = 5, resampling = T, adaptive = T))
```


# SMC tests uniform 


```{r}
likelihood <- function(x) dnorm(x, log = T)
prior <- createUniformPrior(0,5)

bayesianSetup <- createBayesianSetup(likelihood, prior, parallel = T)
finalParticles <- smcSampler(bayesianSetup, initialParticles = 10000, iterations = 10, resampling = T, parallel = T)

hist(finalParticles$particles, breaks = 1000, freq = F, xlim = c(0,5))
library(msm)
curve(dtnorm(x, lower=0, upper = 5), from = 0, to = 5, add = T, col = "red")
```



```{r}
likelihood <- function(x) dnorm(x, log = T)
prior <- createUniformPrior(-10,10)

bayesianSetup <- createBayesianSetup(likelihood, prior, parallel = T)
finalParticles <- smcSampler(bayesianSetup, initialParticles = 10000, iterations = 10, resampling = T, parallel = T)

hist(finalParticles$particles, breaks = 1000, freq = F, xlim = c(-5,5))
curve(dnorm, from = -10, to = 10, add = T, col = "red")

mean(finalParticles$particles)
sd(finalParticles$particles)
```




```{r}
likelihood <- function(x) sum(dunif(x, log = T))
prior <- createUniformPrior(c(0,0),c(5,5))

bayesianSetup <- createBayesianSetup(likelihood, prior, parallel = T)
finalParticles <- smcSampler(bayesianSetup, initialParticles = 10000, iterations = 10, resampling = T, parallel = T)

correlationPlot(finalParticles)
```



# Linear Regression 


First, we define the problem:

```{r}
trueA <- 5
trueB <- 2
trueSd <- 10
sampleSize <- 50
 
# create independent x-values 
x <- (-(sampleSize-1)/2):((sampleSize-1)/2)
# create dependent values according to ax + b + N(0,sd)
y <-  trueA * x + trueB + rnorm(n=sampleSize,mean=0,sd=trueSd)
```

Prior

```{r}
prior <- createUniformPrior(c(0,-10,0),c(10,5,30))
```

Likelihood function:

```{r}
likelihood <- function(param){
    a = param[1]
    b = param[2]
    sd = param[3]
     
    pred = a*x + b
    singlelikelihoods = dnorm(y, mean = pred, sd = sd, log = T)
    sumll = sum(singlelikelihoods)
    return(sumll)   
}
```

And we can call the SMC function:

```{r}
bayesianSetup <- createBayesianSetup(likelihood, prior, parallel = F)

bayesianSetup$likelihood$density(c(1,3,5))

finalParticles <- smcSampler(bayesianSetup, initialParticles = 10000, iterations = 10, resampling = T)
correlationPlot(finalParticles)
```













