---
title: "Test of scaling and mixing in high dimensions for DREAM"
author: "Stefan Paul"
date: "3. Juni 2016"
output: html_document
---

This document contains a test example for the DREAM sampler.
The sampler will be run for multivariate normal distributions with a
growing number of dimensions.
Starting with three dimensions, the sampler will be tested for 
5, 10, 20, 50, 100, 200, 500 and 1000 dimensions.

First a function is defined that takes a number of dimensions and
creates the likelihood function for us.

```{r}
generateTest <- function(mean = NULL, ndim  = 3){
  
  sigma = diag(rep(1,ndim))
  if(is.null(mean)) mean <- sample(1:1000,ndim)
  
  out <- function(x) mvtnorm::dmvnorm(x, mean = mean, sigma = sigma, log=T)
  
  return(out)
}
```

Second the function is applied and the results are plotted for 
the two first parameters.

```{r}
library(BayesianTools)
parVec <- c(3,5,10,20,50,100,200)

settings <- list(iterations = 100000)

count <- 1
test <- list()

for(i in parVec){

mvec <- sample(1:2000,i , replace = F) # Vector of means
lower <- mvec-10 # lower and upper for bayesianSetup
upper <- mvec+10

ll <- generateTest(mean = mvec, ndim = i) # Generate likelihood

# Create Bayesian Setup
BS<- createBayesianSetup(likelihood = ll, lower = lower, upper =upper)

# Run the sampler; Wrap up in capture.output because cat() is not
# captured by KnitR
capture.output(suppressMessages(test[[count]] <- runMCMC(BS, sampler = "DREAM", settings = settings)))
count <- count +1
}

```


```{r}
acc.rate <- numeric()

for(i in 1:(count-1)){
  acc.rate[i] <- test[[i]]$acceptanceRate
  
}


plot(parVec, acc.rate, type = "b", main = "Acceptance rate for different dimensions, DREAM",
     xlab = "Number of parameters", ylab = " Acceptance rate") 

```

```{r}
par(mfrow = c(2,2))
for(i in 1:(count-1)){

  
plot(as.numeric(test[[i]]$chain[[1]][,1]), type = "l", main = paste("Par1 of ", parVec[i]))  
plot(density(test[[i]]$chain[[1]][,1]), type = "l", main = paste("Density of Par1 of ", parVec[i]))  

plot(as.numeric(test[[i]]$chain[[1]][,2]), type = "l", main = paste("Par2 of ", parVec[i]))  
plot(density(test[[i]]$chain[[1]][,2]), type = "l", main = paste("Density of Par2 of ", parVec[i]))  


}


```


Now the same for DREAMzs
```{r}
library(BayesianTools)
parVec <- c(3,5,10,20,50,100,200)

settings <- list(iterations = 100000)

count <- 1
test <- list()

for(i in parVec){

mvec <- sample(1:2000,i , replace = F) # Vector of means
lower <- mvec-10 # lower and upper for bayesianSetup
upper <- mvec+10

ll <- generateTest(mean = mvec, ndim = i) # Generate likelihood

# Create Bayesian Setup
BS<- createBayesianSetup(likelihood = ll, lower = lower, upper =upper)

# Run the sampler; Wrap up in capture.output because cat() is not
# captured by KnitR
capture.output(suppressMessages(test[[count]] <- runMCMC(BS, sampler = "DREAMzs", settings = settings)))
count <- count +1
}

```


```{r}
acc.rate <- numeric()

for(i in 1:(count-1)){
  acc.rate[i] <- test[[i]]$acceptanceRate
  
}


plot(parVec, acc.rate, type = "b", main = "Acceptance rate for different dimensions, DREAMzs",
     xlab = "Number of parameters", ylab = " Acceptance rate") 

```

```{r}
par(mfrow = c(2,2))
for(i in 1:(count-1)){

  
plot(as.numeric(test[[i]]$chain[[1]][,1]), type = "l", main = paste("Par1 of ", parVec[i]))  
plot(density(test[[i]]$chain[[1]][,1]), type = "l", main = paste("Density of Par1 of ", parVec[i]))  

plot(as.numeric(test[[i]]$chain[[1]][,2]), type = "l", main = paste("Par2 of ", parVec[i]))  
plot(density(test[[i]]$chain[[1]][,2]), type = "l", main = paste("Density of Par2 of ", parVec[i]))  


}


```

