---
title: "Estimation of a random effect model with the BT package"
author: "Florian Hartig"
date: "10 Jun 2016"
output: 
  html_document:
    keep_md: yes
    toc: yes
abstract: "This example shows how to specify a random effect model with the BT package" 
editor_options: 
  chunk_output_type: console
---

```{r}
set.seed(123)
library(BayesianTools)

```


# Creation of test case

I create a test dataset with a linear dependency between x and y, with an additional effect that is applied on a grouping factor - you can think of this as a region or site. The structure conforms to a random intercept model.

```{r, fig.width=5, fig.height=5}
a <- 5
b <- 10
sigma <- 10
rsigma = 30
group = rep(1:11, each = 5)
randomEffect = rnorm(11, sd = rsigma)

x <- -27:27
y <- a * x + b + rnorm(55,0,sd = sigma) + randomEffect[group]
plot(x,y, col = group, pch = 3)
```

# Fitting the random effects model with BayesianTools 

```{r}
likelihood <- function(par){
  
  ax = par[1]
  bx = par[2]
  sigmax <- par[3]
  rsigmax = par[4]
  randomEffectx = par[5:15]
  
  llRandom = sum(dnorm(randomEffectx, sd = rsigmax, log = T))
  
  llObservation = sum(dnorm(ax * x + bx + randomEffectx[group] - y , sd = sigmax, log = T))
  return(llRandom + llObservation)
  
}
```


```{r, message=F}
library(BayesianTools)

setup <- createBayesianSetup(likelihood = likelihood, lower = c(-20,-20,0,0,rep(-100,11)), upper = c(20,20,50,100,rep(100,11)))
settings <- list(iterations = 10000, message=F)

res <- runMCMC(bayesianSetup = setup, settings = settings)
summary(res)
marginalPlot(res, which = 1:3, start = 300)

```


# Fitting the model with alternative options

## Bayesian fit with JAGS

```{r}
  # 1) Model definition exactly how we created our data 
  modelCode = "
    model{
      
      # Likelihood
      for(i in 1:i.max){
        y[i] ~ dnorm(mu[i],tau)
        mu[i] <- a*x[i] + b + r[group[i]]
      }

      # random effect
      for(i in 1:nGroups){
        r[i] ~ dnorm(0,rTau)
      }

      # Prior distributions
      a ~ dnorm(0,0.001)
      b ~ dnorm(0,0.001)

      tau <- 1/(sigma*sigma)
      sigma ~ dunif(0,100)

      rTau <- 1/(rSigma*rSigma)
      rSigma ~ dunif(0,100)
    }
  "
  
  # 2) Set up a list that contains all the necessary data (here, including parameters of the prior distribution)
  Data = list(y = y, x = x, i.max = length(y), group = group, nGroups = 11)

  # 3) Specify a function to generate inital values for the parameters
  inits.fn <- function() list(a = rnorm(1), b = rnorm(1), sigma = runif(1,1,100), rSigma = runif(1,1,100))

```


For random effects and similar, I often prefer the R2jags package - not only that it has a bit simpler interface, but it also has different standard plots (those are taken from the STAN / Rstan package), which are a bit more complicated at first, but make it easier to display random effects. 

```{r,  fig.width=18, fig.height=18}
library(R2jags)
system.time(R2JagsResults <- jags(data=Data, inits=inits.fn, parameters.to.save=c("a","b","sigma", "rSigma", "r"), n.chains=1, n.iter=10000, model.file=textConnection(modelCode)))

plot(R2JagsResults)
print(R2JagsResults)

```


## Fitting the model with Maximum Likelihood, lme4

Adding the random intercept on group. Note that the variances are not correctly estimated. 

```{r, fig.width=5, fig.height=5}
library(lme4)
fit <- lmer(y ~ x + (1|group))
summary(fit)
plot(x,y, col = group,  pch = 3)
for(i in 1:11){
  abline(coef(fit)$group[i,1], coef(fit)$group[i,2], col = i)
}

```
