---
title: "MCMC sampler test"
author: "Stefan Paul"
date: "9. Juni 2016"
output: html_document
---
Test file for all sampler in the BayesianTools package.
The sampler are tested for the following cases: 


* 3d normal distribution 
* 3d normal distribution with truncated likelihood
* 3d normal distribution with truncated prior 
* 30d normal distribution
* 50d normal distribution


```{r}
library(BayesianTools)
library(truncnorm)
``` 

First a function for the test cases  is created:
```{r}
generateTest <- function(mean = NULL){
  ndim <- length(mean)
  sigma = diag(rep(1,ndim))
  out <- function(x) mvtnorm::dmvnorm(x, mean = mean, sigma = sigma, log=T)
  return(out)
}
```

And a function that is running all the samplers.
```{r}

runSampler <- function(bayesianSetup, iterations = 50000){
  
  BS <- bayesianSetup
  
  settings <- list(iterations = iterations, adapt = F, optimize = F, message = FALSE)
  testMH <- runMCMC(BS, sampler = "Metropolis", settings = settings)

  settings <- list(iterations = iterations, adapt = T, optimize = F, message = FALSE)
  testAM <- runMCMC(BS, sampler = "Metropolis", settings = settings)

  settings <- list(iterations = iterations, adapt = T, optimize = F, DRlevels = 2, message = FALSE)
  testDRAM <- runMCMC(BS, sampler = "Metropolis", settings = settings)

  settings <- list(iterations = iterations, message = FALSE)
  testDE <- runMCMC(BS, sampler = "DE", settings = settings)
  
  settings <- list(iterations = iterations, blockUpdate = list("correlation", h = 0.5), message = FALSE)
  testDEblock <- runMCMC(BS, sampler = "DE", settings = settings)

  settings <- list(iterations = iterations, message = FALSE)
  testDEzs <- runMCMC(BS, sampler = "DEzs", settings = settings)
  
  settings <- list(iterations = iterations, blockUpdate = list("correlation", h = 0.5), message = FALSE)
  testDEzsblock <- runMCMC(BS, sampler = "DEzs", settings = settings)
  
  settings <- list(iterations = iterations, message = FALSE)
  testDREAM <- runMCMC(BS, sampler = "DREAM", settings = settings)

  settings <- list(iterations = iterations, message = FALSE)
  testDREAMzs <- runMCMC(BS, sampler = "DREAMzs", settings = settings)
  
  settings <- list(iterations = iterations, message = FALSE)
  testTwalk <- runMCMC(BS, sampler = "Twalk", settings = settings)
  
  out <- list("MH" = testMH, "AM" = testAM, "DRAM" = testDRAM, "DE" = testDE,
              "DEzs" = testDEzs, "DEblock" = testDEblock, "DEzsblock" = testDEzsblock,
              "DREAM" = testDREAM, "DREAMZS" = testDREAMzs, "Twalk" = testTwalk)
  
  return(out)
  
}
```


# 3d normal distibution
```{r, message = FALSE}

ll <- generateTest(mean = c(0,2,4))
BS <- createBayesianSetup(likelihood = ll, lower = rep(-10, 3), upper = rep(10, 3),
                                          parallel = FALSE)

Normal3d <- runSampler(BS)
```

### Plot the results
```{r}
x <- seq(-11,11, len = 100)

for(i in 1:length(Normal3d)){

  if(Normal3d[[i]]$settings$sampler == "Metropolis"){
  plot(density(Normal3d[[i]]$chain[,1]), ylim = c(0,1),xlim = range(x), main = names(Normal3d)[i])
  }else plot(density(Normal3d[[i]]$chain[[1]][,1]), ylim = c(0,1),xlim = range(x), main = names(Normal3d)[i])
  lines(x,dtruncnorm(x, a = -10, b = 10), col = "red", type = "l", lwd = 2)
}


```




# Truncated prior.
```{r}

ll <- generateTest(mean = c(0,10,12))
BS <- createBayesianSetup(likelihood = ll, lower = c(0,-10,-10), upper = rep(30, 3),
                                          parallel = FALSE)

TruncatedPrior <- runSampler(BS)
```

### Plot the results
```{r}
x <- seq(0,10, len = 100)

for(i in 1:length(TruncatedPrior)){

  if(TruncatedPrior[[i]]$settings$sampler == "Metropolis"){
  plot(density(TruncatedPrior[[i]]$chain[,1]), ylim = c(0,1),xlim = range(x), main = names(TruncatedPrior)[i])
  }else plot(density(TruncatedPrior[[i]]$chain[[1]][,1]), ylim = c(0,1), xlim = range(x),main = names(TruncatedPrior)[i])
  lines(x,dtruncnorm(x, a = 0, b = 30), col = "red", type = "l", lwd = 2)
}

```


# Truncated likelihood
```{r}
generateTestTrunc <- function(mean = NULL){
  ndim <- length(mean)
  sigma = diag(rep(1,ndim))
  out <- function(x) {
    if(x[1]<0) return(-Inf)
    else mvtnorm::dmvnorm(x, mean = mean, sigma = sigma, log=T)
  }
  return(out)
}

ll <- generateTestTrunc(mean = c(0,2,4))
BS <- createBayesianSetup(likelihood = ll, lower = rep(-10, 3), upper = rep(10, 3),
                                          parallel = FALSE)

TruncatedLL <- runSampler(BS)

```


### Plot the results
```{r}
x <- seq(0,10, len = 100)

for(i in 1:length(TruncatedLL)){

  if(TruncatedPrior[[i]]$settings$sampler == "Metropolis"){
  plot(hist(TruncatedPrior[[i]]$chain[,1]), ylim = c(0,1), xlim = range(x),main = names(TruncatedPrior)[i], freq = FALSE)
  }else plot(hist(TruncatedPrior[[i]]$chain[[1]][,1]), ylim = c(0,1), xlim = range(x), main = names(TruncatedPrior)[i], freq = FALSE)
  lines(x,dtruncnorm(x, a = 0, b = 10), col = "red", type = "l", lwd = 2)
}

```



# 30d normal distibution
```{r, message = FALSE}

meanvec <- sample(1000, 30)
ll <- generateTest(mean =meanvec)
BS <- createBayesianSetup(likelihood = ll, lower = meanvec-10, upper = meanvec+10,
                                          parallel = FALSE)

out <-capture.output(suppressMessages(Normal30d <- runSampler(BS)))
```

### Plot the results
```{r}
x <- seq(meanvec[1]-10,meanvec[1]+10, len = 100)

for(i in 1:length(Normal30d)){

  if(Normal30d[[i]]$settings$sampler == "Metropolis"){
  plot(density(Normal30d[[i]]$chain[,1]), ylim = c(0,1),xlim = range(x), main = names(Normal30d)[i], freq = FALSE)
  }else plot(density(Normal30d[[i]]$chain[[1]][,1]), ylim = c(0,1),xlim = range(x), main = names(Normal30d)[i], freq = FALSE)
  lines(x,dtruncnorm(x, a = meanvec[1]-10, b = meanvec[1]+10, mean = as.numeric(meanvec[1])), col = "red", type = "l", lwd = 2)
}


```



# 50d normal distibution
```{r, message = FALSE}

meanvec <- sample(1000, 50)
ll <- generateTest(mean =meanvec)
BS <- createBayesianSetup(likelihood = ll, lower = meanvec-10, upper = meanvec+10,
                                          parallel = FALSE)

Normal50d <- runSampler(BS)
```

### Plot the results
```{r}
x <- seq(meanvec[1]-10,meanvec[1]+10, len = 100)

for(i in 1:length(Normal50d)){

  if(Normal50d[[i]]$sampler == "Metropolis"){
  plot(hist(Normal50d[[i]]$chain[,1]), ylim = c(0,1),xlim = range(x), main = names(Normal50d)[i], freq = FALSE)
  }else plot(hist(Normal50d[[i]]$chain[[1]][,1]), ylim = c(0,1), xlim = range(x),main = names(Normal50d)[i], freq = FALSE)
  lines(x,dtruncnorm(x, a = meanvec[1]-10, b = meanvec[1]+10, mean = as.numeric(meanvec[1])), col = "red", type = "l", lwd = 2)
}


```
