---
title: "Supplementary material - priors and estimated referece posterior distributions"
output: 
  html_document: 
    keep_md: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = F, fig.align="center", fig.width = 8)
```

```{r, echo = F,  message = F}
library(BayesianTools)
set.seed(123)
```

# Prior information 

## VSEM

The VSEM model is implemented in the BayesianTools package. A model description, including a detailed description of the meaning of the parameters can be found via ?BayesianTools::VSEM. 


The following reference parameters (for the creation of the synthetic data) and prior ranges were used for the example with strong posterior parameter correlation (equifinality). 

```{r, echo = F}
refPars <- VSEMgetDefaults()
# this adds one additional parameter for the likelihood standard deviation (see below)
refPars[12,] <- c(2, 0.1, 4) 
rownames(refPars)[12] <- "error-sd"
parSel = c(1:6, 12)
refPars[parSel,]
```


The following reference parameters (for the creation of the synthetic data) and prior ranges were used for the example with weak posterior parameter correlation (no equifinality).

```{r, echo = F}
parSel <- c(1:3, 5:7, 12)
refPars[parSel,]
```


## 3PG

For reproducing our resutls, you should install the 3PG implementation in the threePGN package available via https://github.com/ForModLabUHel/threePGN-package

For new simulations, we recommend switching to the r3PG package, which is available from CRAN and described in Trotsiuk, Volodymyr, Florian Hartig, and David I. Forrester. "r3PG–An r package for simulating forest growth using the 3‐PG process‐based model." Methods in Ecology and Evolution 11.11 (2020): 1470-1475. This package and the paper also include some extended documentation of the model. 

The following reference parameters (for the creation of the synthetic data) and prior ranges were used for the 3PG model. 

```{r, echo = F}
library(threePGN)
ref.pars <- c(data_param$mode, 1, 1)
min.pars <- c(data_param$min, 0.2, 0.2)
max.pars <- c(data_param$max, 2, 2)
parameters = data.frame(reference = ref.pars, priorMin = min.pars, priorMax = max.pars)
parameters
```

The meaning and units of the parameters are described in 

```{r, eval = F}
?threePGN::data_param
```


# Reference posterior estimates

Here, we show the reference posterior estimates, created with an extremely long DEzs sampler. It is assumed that these are very close to the "true" (= mathematically exact) posterior distributions, and can thus be used as a reference. Since the model used in the 3PGN_sleep experiments is identical to the one in 3PGN, figures for the 3PGN_sleep are omitted. 

The SMC posterior estimates for the different settings are avaialble [here](https://www.dropbox.com/s/dndmyv8atg8vhsy/SMCruns.html?dl=0). To reproduce the underlying scripts, the MCMC / SMC samplers need to be placed in a subfolder data of this folder. 

## VSEM - Case with strong correlation

Here the marginal posterior uncertainty

```{r , echo=FALSE, fig.height=8}
load("./data/reference_MCMC/reference_VSEMa_list.RData")
marginalPlot(save.list[[3]], singlePanel = T)
```

Here the posterior parameter correlation 

```{r, fig.height=8}
correlationPlot(save.list[[3]])
```


## VSEM - Case without strong correlation

Here the marginal posterior uncertainty

```{r , echo=FALSE, fig.height=8}
load("./data/reference_MCMC/reference_VSEMb_list.RData")
marginalPlot(save.list[[3]], singlePanel = T)
```

Here the posterior parameter correlation 

```{r, echo=FALSE, fig.height=8}
correlationPlot(save.list[[3]])
```


## 3PG

Here the marginal posterior uncertainty

```{r , echo=FALSE, fig.height=40}
load("./data/reference_MCMC/reference_threePGN_list.RData")
marginalPlot(save.list[[3]], singlePanel = T)
```

Here the posterior parameter correlation, for reasons of space only for the parameters with the visually largest marginal posterior uncertainty

```{r, echo=FALSE, fig.height=13, fig.width=13}
correlationPlot(save.list[[3]], whichParameters = c(1,5,6,13,14,20,21, 23,26,30,47,52,53))
```
