---
title: "Practical examples fitting models with Bayesian Tools"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Practical examples fitting models with Bayesian Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
abstract: "This vignette shows how to fit standard statistical models with the BayesianTools package"  
author: Florian Hartig
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=7, warning=FALSE, cache = F)
```

```{r, echo = F, message = F}
set.seed(123)
```


# Linear regression

Consider the following linear regression model

```{r}
fit = lm(Ozone ~ Wind, data = airquality)
summary(fit)
```

How to do the same with the BT package? First, we need to specify the likelihood for a linear regression:

```{r}
dat = model.frame(fit) # using model.frame makes sure we use the same data

likelihood <- function(par){
  # c("a", "b", "err-sd")
  pred = par[1] * dat$Wind +  par[2]
  ll = sum(dnorm(pred - dat$Ozone, 
                            sd = par[3], 
                            log = T))
  return(ll)
}

par = c(-5,96,25)
likelihood(par)
```

Creating the BayesianSetup by combining likelihood with uniform (flat) priors

```{r}
library(BayesianTools)
setup <- createBayesianSetup(likelihood = likelihood, 
                             lower = c(-100,-100,0), 
                             upper = c(100,200,50),
                             names = c("Wind", "intercept", "err-sd"))
```

Creating required setup for wrapper function runMCMC to start MCMCs

```{r, message=F}
settings <- list(iterations = 10000, nrChains = 2, message = F)
res <- runMCMC(bayesianSetup = setup, settings = settings)
```

Traceplots and Gelman diagnostics to check convergence confirm the estimation procedure is fine and the MCMC has converged.

```{r}
plot(res)
gelmanDiagnostics(res)
```

Summary table is virtually identical to the standard regression table using lm 

```{r}
summary(res, start=1000)
```

Plots of the marginal prior and posterior distribution

```{r}
marginalPlot(res, start = 1000, prior = T)
```

A correlation plot shows a very strong correlction between slope and intercept. This is because the predictor variable wasn't centered. 

```{r}
correlationPlot(res, start = 1000)
```

Tip: in a Bayesian analysis, it is useful to scale and center all predictors. Scaling helps to set sensible non-informative priors. Centering removes posterior correlations between variables. 

# Linear mixed model with the BT package

To demonstrate how to fit a linera mixed model, I create a test dataset with a linear dependency between x and y, with an additional effect that is applied on a grouping factor - you can think of this as a region or site. The structure conforms to a random intercept model.

```{r, fig.width=5, fig.height=5}
a <- 5
b <- 10
sigma <- 10
rsigma = 30
group = rep(1:11, each = 5)
randomEffect = rnorm(11, sd = rsigma)

x <- -27:27
y <- a * x + b + rnorm(55,0,sd = sigma) + randomEffect[group]
plot(x,y, col = group, pch = 3)
```

With lme4, we would fit the model as 

```{r, eval = F}
library(lme4)
fit <- lmer(y ~ x + (1|group))
```

Fitting the random effects model with BayesianTools 

```{r}
likelihood <- function(par){
  # c("a", "b", "err-sd", "RE-sd", 1:11)
  llObservation = sum(dnorm(par[1] * x +  par[2] + par[group+4] - y , sd = par[3], log = T))
  llRandom = sum(dnorm(par[5:15], sd = par[4], log = T))
  return(llRandom + llObservation)
}

likelihood(c(4,10,6,20,rep(0,11)))
```

Estimating the model

```{r}
library(BayesianTools)
setup <- createBayesianSetup(likelihood = likelihood, 
                             lower = c(-20,-50,0,0,rep(-100,11)), 
                             upper = c(20,50,50,100,rep(100,11)),
                             names = c("a", "b", "err-sd", "RE-sd", 1:11))
settings <- list(iterations = 30000, nrChains = 2, message = F)

res <- runMCMC(bayesianSetup = setup, settings = settings)
```

Traceplots for the 4 main parameters

```{r}
plot(res, which = 1:4)
```

Looks good, let's look at the summary

```{r}
summary(res, start=5000)
```


```{r}
marginalPlot(res, start = 5000, prior = T, which = 1:4)
```

Comparison with STAN fits via brms 

```{r, eval = F}
library(brms)
fit <- brm(y ~ x + (1|group), data = data.frame(x=x, y=y, group = group))
summary(fit)
```





