---
title: "Test"
output: html_document
date: "2023-09-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Linear regression with the BT package



```{r}
fit = lm(Ozone ~ Wind, data = airquality)
summary(fit)
```

```{r}

dat = airquality[complete.cases(airquality), ]

# Fitting the random effects model with BayesianTools 

```{r}
likelihood <- function(par){
  # c("a", "b", "err-sd")
  llObservation = sum(dnorm(par[1] * dat$Wind +  par[2] - dat$Ozone, sd = par[3], log = T))
  return(llObservation)
}

likelihood(c(4,10,6))
```



```{r, message=F}
library(BayesianTools)
setup <- createBayesianSetup(likelihood = likelihood, 
                             lower = c(-100,-100,0), 
                             upper = c(100,200,50),
                             names = c("a", "b", "err-sd"))
settings <- list(iterations = 10000, nrChains = 2)

res <- runMCMC(bayesianSetup = setup, settings = settings)
plot(res)
marginalPlot(res, start = 1000, prior = T)
summary(res, start=1000)
gelmanDiagnostics(res)
```


# Linear mixed model with the BT package





# Fitting a ecosystem model with the BT package 


?VSEM   
How do I know lower and upper?


