---
title: "Test"
output: html_document
date: "2023-09-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this vignette, we want to illustrate how BayesianTools can be implemented on a real data.

In this case, we are using airquality data.

For the illustration purpose, we are fitting a linear regression model on the data - so that we can compare it with our result.


```{r}
fit = lm(Ozone ~ Wind, data = airquality)
summary(fit)
```

```{r}

dat = airquality[complete.cases(airquality), ]

```

Creating a likelihood function from the data.

```{r}
likelihood <- function(par){
  # c("a", "b", "err-sd")
  llObservation = sum(dnorm(par[1] * dat$Wind +  par[2] - dat$Ozone, sd = par[3], log = T))
  return(llObservation)
}

likelihood(c(4,10,6))
```

Creating a standardized collection of prior, likelihood and posterior function.

```{r, message=F}
library(BayesianTools)
setup <- createBayesianSetup(likelihood = likelihood, 
                             lower = c(-100,-100,0), 
                             upper = c(100,200,50),
                             names = c("a", "b", "err-sd"))
```

Creating required setup for wrapper function runMCMC to start MCMCs

```{r}
settings <- list(iterations = 10000, nrChains = 2)
res <- runMCMC(bayesianSetup = setup, settings = settings)

summary(res, start=1000)
```


```{r}
plot(res)
```

```{r}
marginalPlot(res, start = 1000, prior = T)
```

```{r}
gelmanDiagnostics(res)
```
```{r}
correlationPlot(res)
```

# Linear mixed model with the BT package





# Fitting a ecosystem model with the BT package 


?VSEM   
How do I know lower and upper?


