---
title: "Fitting standard statistical models with the BayesianTools Package"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Statistical Models with the BayesianTools Package}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
abstract: "This vignette shows how to fit standard statistical models with the BayesianTools package" 
author: Florian Hartig
editor_options: 
  chunk_output_type: console
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=5, warning=FALSE, cache = F)
```

```{r, echo = F, message = F}
set.seed(123)
```


# Linear regression

Consider the following linear regression model

```{r}
fit = lm(Ozone ~ Wind, data = airquality)
summary(fit)
```

How to do the same with the BT package? First, we need to specify the likelihood for a linear regression:

```{r}
dat = model.frame(fit) # using model.frame makes sure we use the same data

likelihood <- function(par){
  # c("a", "b", "err-sd")
  pred = par[1] * dat$Wind +  par[2]
  ll = sum(dnorm(pred - dat$Ozone, 
                            sd = par[3], 
                            log = T))
  return(ll)
}

par = c(-5,96,25)
likelihood(par)
```

Creating the BayesianSetup by combining likelihood with uniform (flat) priors

```{r, message=F}
library(BayesianTools)
setup <- createBayesianSetup(likelihood = likelihood, 
                             lower = c(-100,-100,0), 
                             upper = c(100,200,50),
                             names = c("Wind", "intercept", "err-sd"))
```

Creating required setup for wrapper function runMCMC to start MCMCs

```{r}
settings <- list(iterations = 10000, nrChains = 2)
res <- runMCMC(bayesianSetup = setup, settings = settings)
```

Traceplots and Gelman diagnostics to check convergence confirm the estimation procedure is fine and the MCMC has converged.

```{r}
plot(res)
gelmanDiagnostics(res)
```

Summary table is virtually identical to the standard regression table using lm 

```{r}
summary(res, start=1000)
```

Plots of the marginal prior and posterior distribution

```{r}
marginalPlot(res, start = 1000, prior = T)
```

A correlation plot shows a very strong correlction between slope and intercept. This is because the predictor variable wasn't centered. 

```{r}
correlationPlot(res, start = 1000)
```

Tip: in a Bayesian analysis, it is useful to scale and center all predictors. Scaling helps to set sensible non-informative priors. Centering removes posterior correlations between variables. 

# Linear mixed model with the BT package





# Fitting a ecosystem model with the BT package 



